Exploring probability distributions for bivariate temporal granularities

### 2019-12-09 main.Rmd
visualization section 1 naming and corresponding changes

## Choice of statistical distribution plots

@@

There are several ways to plot statistical distributions. The entire distribution can be displayed or some quantiles can be visualized with probabilities in these plots either computed using kernel density estimation methods or quantile estimation methods. The latter do not allow us to see the shape, skewness, nature of the tail or multi-modality with so much clarity as the former. However, they avoid much clutter, where some specific probabilities representing typical and extreme behavior are on focus.  Density plots can show the entire distribution, unlike discretizing and showing only parts of it. The methods are subject to the same sample size restrictions and challenges that apply to any density estimation. In practice, the densities are estimated reasonably with at least 30 observations. Even with sample sizes of several hundred, however, choosing too large a value for bandwidth can cause the estimate to oversmooth the data.

@@

### Quantile based methods

Most commonly used techniques to display a distribution of data include the box-and-whisker plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers. The box plot is a compact distributional summary. Due to wide spread popularity and simplicity in implementation, a number of variations are proposed to the original one which provides alternate definitions of quantiles, whiskers, fences and outliers. Notched box plots [@Mcgill1978-hg] has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots.<!-- Challenge of boxplot --> The standard box plot and all of these variations are helpful to get an idea of the distribution at a glance. However, for data less than 1000 observations, detailed estimates of tail behavior beyond the quartiles are not trustworthy. Also, the number of outliers is large for larger data set since number of outliers is proportional to the number of observations.

<!-- Description and lit review of letter-value plots -->
The letter-value box plot [@Hofmann2017-sg] was designed to adjust for number of outliers proportional to the data size and display more reliable estimates of tail. “outliers” in letter value plots are those observations beyond the most extreme letter value. The letter values are shown till the depths where the letter values are reliable estimates of their corresponding quantiles and hence might lead to a lot of letter values being shown, leading to overload of information in one plot.

@@

Quantile plots visually portray the quantiles of the distribution of sample data. These plots can display any quantiles instead of quartiles in a traditional boxplot. This gives more information than box plots and yet avoid clutter by enabling us to focus just on specific probabilities. When reviewing a boxplot, an outlier is defined as a data point that is located outside the fences (“whiskers”) of the boxplot (e.g. outside 1.5 times the interquartile range above the upper quartile and below the lower quartile). However, outliers are open to interpretation and not shown in an quantile plot.

@@

### Kernel density based methods

>>>>

## Choice of exploratory visualization tool

@@

There are several ways to visualize statistical distributions. The entire distribution can be displayed through density estimates using non-paramteric density estimation methods or only summarized values can be displayed using descriptive statistics. The latter do not allow us to see the shape, skewness, nature of the tail or multi-modality with so much clarity as the former. However, they avoid much clutter, where some specific probabilities representing typical and extreme behavior are on focus. Non-parametric density based plots can show the entire distribution, unlike discretizing and showing only parts of it. These displays depend on the specific estimation procedure (e.g., kernel density estimate) as well as on additional smoothing parameters and hence can vary for the same data set. Furthermore, these methods are subject to the same sample size restrictions and challenges that apply to any density estimation. In practice, the densities are estimated reasonably with at least 30 observations. However, even with large sample size, choosing too large a value for bandwidth can cause the estimate to oversmooth the data. The visualization tools based on descriptive statistics display only actual observations from the sample, and remain free of any tuning parameters or Kernels.

@@

### Descriptive statistics based methods

Most commonly used techniques include the box plots [@Tukey1977-jx] which convey statistical features such as the median, quartile boundaries, hinges, whiskers and extreme outliers. These displays are based on descriptive statistics, for example, $([n/2] + 1)/2^{th}$ and $(n + 1 - ([n/2] + 1)/2)^{th}$ order statistics are taken as estimates of the quartiles. Due to wide spread popularity and simplicity in implementation, a number of variations are proposed to the original one which provides alternate definitions of quantiles, whiskers, fences and outliers. Notched box plots [@Mcgill1978-hg] has box widths proportional to the number of points in the group and display confidence interval around medians aims to overcome some drawbacks of box plots.<!-- Challenge of boxplot --> The standard box plot and all of it's variations are helpful to get an idea of the distribution at a glance. However, for data less than 1000 observations, detailed estimates of tail behavior beyond the quartiles are not trustworthy. Also, the number of outliers is large for larger data set since number of outliers is proportional to the number of observations.

<!-- Description and lit review of letter-value plots -->
The letter-value box plot [@Hofmann2017-sg] was designed to adjust for number of outliers proportional to the data size and display more reliable estimates of tail. These are particularly useful for large data sets. “outliers” in letter value plots are those observations beyond the most extreme letter value. The letter values are shown till the depths where the letter values are reliable estimates of their corresponding quantiles and hence might lead to a lot of letter values being shown, leading to overload of information in one plot.

@@

Quantile plots visually portray the quantiles of the distribution of sample data. These plots can display any quantiles instead of quartiles in a traditional boxplot. Some of the most implemented definitions of sample quantiles based on one or two order statistics can be found in @Hyndman1996-ty. These plots give more information than box plots and yet avoid clutter by enabling us to focus just on specific probabilities. When reviewing a boxplot, an outlier is defined as a data point that is located outside the fences (“whiskers”) of the boxplot (e.g. outside 1.5 times the interquartile range above the upper quartile and below the lower quartile). However, outliers are open to interpretation and not shown in an quantile plot.

@@

### Non-parametric density based methods
